Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-12 submission/manuscript.Rmd
edit results 3

## _Kdm5c_-KO epiblast-like cells aberrantly express master regulators of germline identity

@@

Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014]. A subset of epiblast stem cells become the primordial germ cells (PGCs) while the remainder differentiate into the ectoderm, mesoderm, and endoderm to form the somatic tissues[@bardotMouseGastrulationCoordination2020]. This developmental time point can be modeled _in vitro_ through differentiation of embryonic stem cells (ESCs) into epiblast-like stem cells (EpiLCs) (Figure 3A, top)[@hayashiReconstitutionMouseGerm2011; @samantaGeneratingPrimedPluripotent2020]. While some germline-enriched genes are also expressed in embryonic stem cells (ESCs) and in the 2-cell stage[@wellingDAZLRegulatesTet12015; @macfarlanEmbryonicStemCell2012; @suzukiLossMAXResults2016], they are silenced as they differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021]. Therefore, we tested if KDM5C was necessary for silencing germline genes at this developmental stage by evaluating the impact of _Kdm5c_ loss in EpiLCs.

We first identified _Kdm5c_-KO EpiLC DEGs in our previously published RNA-seq dataset[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1) and assessed tissue-enriched gene expression. Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3B). Primed pluripotency genes, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_, were unaltered in _Kdm5c_-KO EpiLCs(Figure 3C). Furthermore, _Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation (Figure 3D) was normal, indicating KDM5C loss does not impair EpiLC formation.

To determine if germline DEGs are constitutively dysregulated or if they can change over the course of development, we compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain. We found the majority of germline DEGs were unique to either EpiLCs or the brain, with only _Cyct_ shared across all tissue/cell types (Figure 3E-F). We found particularly high enrichment of meiosis-related gene ontologies in EpiLCs (Figure 3G), such as meiotic cell cycle process (GO: 1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO: 0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`). While there was modest enrichment of meiotic gene ontologies in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes, such those involved in the sperm axoneme structure.

Notably, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3H). These genes are typically expressed during embryonic germ cell development to commit PGCs to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019]. Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo[@macfarlanEmbryonicStemCell2012; @guptaGenomewideScreenReveals2023]. However, we did not see misexpression of two-cell embryo-specific genes, like _Duxf3 (Dux)_ (q = `r signif(duxf3, 3)`) and _Zscan4d_ (q = `r signif(zscan4d, 3)`), indicating _Kdm5c_-KO in EpiLCs do not revert back to a 2-cell state (Figure 3H).

## Females have increased sensitivity to germline gene misexpression with _Kdm5c_ loss
```{r, DESeq2_params_XX, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
#get the parameters used repeatedly, like the log2fold change cut off
source("../../KDM5C_Germ_Dev_Sex_RA/code/utilities/parameters.R")
```

@@

In the early embryo, germline gene promoters are initially silenced by repressive histone modifications, including histone 2A lysine 119 monoubiquitination (H2AK119ub1) and histone 3 lysine 9 trimethylation (H3K9me3), and are then silenced long-term via DNA CpG methylation (CpGme)[@endohPCGF6PRC1SuppressesPremature2017; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021]. Our results above suggest KDM5C also acts during this time period, however how KDM5C interacts with other germline gene silencing mechanism is currently unclear. KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007]. However, KDM5C's catalytic activity was recently shown to not be required for suppressing _Dazl_ in undifferentiated ESCs[@guptaGenomewideScreenReveals2023]. Since H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015], KDM5C's catalytic activity may instead be required for CpGme-mediated, long-term silencing of germline genes. In support of this, CpGme is markedly reduced at two germline gene promoters in the _Kdm5c_-KO adult hippocampus[@scandagliaLossKdm5cCauses2017]. Based on these observations, we hypothesized KDM5C erases H3K4me3 to promote the initial placement of CpGme at germline gene promoters in EpiLCs.

>>>>

## _Kdm5c_-KO epiblast-like cells aberrantly express key regulators of germline identity

@@

Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014]. A subset of epiblast stem cells become the primordial germ cells (PGCs) while the remainder differentiate into the ectoderm, mesoderm, and endoderm to form the somatic tissues[@bardotMouseGastrulationCoordination2020]. This developmental time point can be modeled _in vitro_ through differentiation of embryonic stem cells (ESCs) into epiblast-like stem cells (EpiLCs) (Figure 3A, top)[@hayashiReconstitutionMouseGerm2011; @samantaGeneratingPrimedPluripotent2020]. While some germline-enriched genes are also expressed in na&#239;ve embryonic stem cells (ESCs) and in the 2-cell stage[@wellingDAZLRegulatesTet12015; @macfarlanEmbryonicStemCell2012; @suzukiLossMAXResults2016], they are silenced as they differentiate into EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021]. Therefore, we tested if KDM5C was necessary for silencing germline genes at this developmental stage by evaluating the impact of _Kdm5c_ loss in male EpiLCs.

We first identified _Kdm5c_-KO EpiLC DEGs in our previously published RNA-seq dataset[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1) and then assessed tissue-enriched gene expression. Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3B). Markers of primed pluripotency, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_, were unaltered in _Kdm5c_-KO EpiLCs (Figure 3C). Furthermore, _Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation (Figure 3D) appeared normal, indicating KDM5C loss does not impair EpiLC formation.

To determine if germline DEGs are constitutively dysregulated or if they can change over the course of development, we next compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain. We found the majority of germline DEGs were unique to either EpiLCs or the brain, with only _Cyct_ shared across all tissue/cell types (Figure 3E-F). EpiLCs had particularly high enrichment of meiosis-related gene ontologies (Figure 3G), such as meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`). While there was modest enrichment of meiotic gene ontologies in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes, such those involved in sperm axoneme assembly (GO:0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`) and sperm motility (GO:0097722, p.adjust = `r signif(getGO("sperm motility", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`).

Notably, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3H). These genes are typically expressed during embryonic germ cell development to commit PGCs to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019]. Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo[@macfarlanEmbryonicStemCell2012; @guptaGenomewideScreenReveals2023]. However, we did not see misexpression of two-cell embryo-specific genes, like _Duxf3 (Dux)_ (q = `r signif(duxf3, 3)`) and _Zscan4d_ (q = `r signif(zscan4d, 3)`), indicating _Kdm5c_-KO in EpiLCs do not revert back to a 2-cell-like state (Figure 3H). Altogether, these results demonstrate that while the _Kdm5c_-KO brain primarily expresses spermiogenesis genes, _Kdm5c_-KO EpiLCs express key drivers of germline identity and meiosis.

## Female epiblast-like cells have increased sensitivity to germline gene misexpression with _Kdm5c_ loss
```{r, DESeq2_params_XX, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
#get the parameters used repeatedly, like the log2fold change cut off
source("../../KDM5C_Germ_Dev_Sex_RA/code/utilities/parameters.R")
```

@@

In the early embryo, germline gene promoters are initially silenced by repressive histone modifications, including histone 2A lysine 119 monoubiquitination (H2AK119ub1) and histone 3 lysine 9 trimethylation (H3K9me3), and are then silenced long-term via DNA CpG methylation (CpGme)[@borgelTargetsDynamicsPromoter2010; @endohPCGF6PRC1SuppressesPremature2017; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021]. Our results above suggest KDM5C also acts during this time period, however how KDM5C interacts with other germline gene silencing mechanism is currently unclear. KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007]. However, KDM5C's catalytic activity was recently shown to not be required for suppressing _Dazl_ in undifferentiated ESCs[@guptaGenomewideScreenReveals2023]. Since H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015], KDM5C's catalytic activity may instead be required for CpGme-mediated, long-term silencing of germline genes. In support of this, CpGme is markedly reduced at two germline gene promoters in the _Kdm5c_-KO adult hippocampus[@scandagliaLossKdm5cCauses2017]. Based on these observations, we hypothesized KDM5C erases H3K4me3 to promote the initial placement of CpGme at germline gene promoters in EpiLCs.
